---
title: "Audit Checklist (For Maintainers)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Audit Checklist (For Maintainers)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Purpose

This checklist is for **package maintainers and peer reviewers** who want to verify that `fury` meets quality standards.

**For regular users:** See the [Complete Beginner's Guide](novice-walkthrough.html) instead.

---

# Quick Start

## Run the Automated Audit

```{bash, eval=FALSE}
# From package root directory
Rscript tools/run_audit.R
```

**Expected output:**
```
✅ AUDIT PASSED
All automated checks passed.
```

## Run the Test Suite

```{r, eval=FALSE}
devtools::test()
```

---

# Full Documentation

For the complete 28-point audit checklist, see:

**`inst/audit/AUDIT_CHECKLIST.md`**

This document contains:

- Detailed PASS/FAIL criteria for each check
- Commands to run for manual verification
- Expected artifacts and failure conditions
- Six audit sections:
  1. Novice Walkthrough Audit
  2. SPSS .sav Ingestion Audit
  3. Partitioning + Screening + CONSORT Audit
  4. Methods Write-Up Readiness Audit
  5. Silent Misuse Audit
  6. Determinism Audit

---

# What the Audit Verifies

✅ **Novice usability** — Package is usable for researchers new to R

✅ **SPSS ingestion** — `.sav` files can be imported with tutorials

✅ **Conservative language** — No inflated claims ("final dataset", "cleaned data")

✅ **Methods artifacts** — All required files for peer review are generated

✅ **Determinism** — Same inputs → same outputs

✅ **Governance compliance** — Ecosystem contracts are obeyed

---

# Quick Reference

| Check Category | Automated? | Location |
|----------------|------------|----------|
| README quality | ✅ Yes | `tools/run_audit.R` |
| SPSS fixtures | ✅ Yes | `tools/run_audit.R` |
| Artifact generation | ✅ Yes | `tools/run_audit.R` |
| Conservative language | ✅ Yes | `tools/run_audit.R` + `tests/` |
| Determinism | ✅ Yes | `tools/run_audit.R` + `tests/` |
| Tutorial completeness | ⚠️ Partial | Manual review required |

---

**For full details:** See `inst/audit/AUDIT_CHECKLIST.md` in the package source.
