---
title: "Complete Beginner's Guide to fury"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Beginner's Guide to fury}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Welcome! This Guide is For You If...

- ✅ You're new to R (or have only used it a little bit)
- ✅ You collected data in Qualtrics and exported an SPSS file
- ✅ You need to document which participants you excluded and why
- ✅ You're preparing a manuscript for journal submission
- ✅ You want a clear audit trail for peer reviewers

**You do NOT need to be an R expert to use this guide.**

---

# What is fury?

`fury` is an R package that helps you:

1. **Import your data** from Qualtrics (SPSS format)
2. **Apply screening rules** (like excluding people who didn't consent)
3. **Track what you did** in files that reviewers can check
4. **Prepare your dataset** for analysis

**What fury does NOT do:**

- ❌ Does NOT run statistical tests (t-tests, ANOVA, regression, etc.)
- ❌ Does NOT compute scale scores or reverse-code items
- ❌ Does NOT check reliability (Cronbach's alpha, factor analysis, etc.)
- ❌ Does NOT create APA-formatted tables or figures

Think of `fury` as the **data preparation** step. You'll use other tools (or other R packages) for the actual analysis.

---

# Key Concepts You Need to Understand

## 🔑 Concept 1: Flagging vs. Excluding (VERY IMPORTANT!)

This is the **#1 mistake** beginners make with `fury`.

### Flagging = Marking (Participants Stay In Your Data)

When you **flag** someone, you're putting a note on their data that says "hey, this person failed an attention check" BUT **they are still in your dataset**.

**Example:**
```
Participant 42 failed the attention check
→ fury adds a column: attn_check_failed = TRUE
→ Participant 42 is STILL in your data
→ You decide later whether to remove them
```

**Use flagging when:** You want to keep participants for sensitivity analysis (testing whether results change if you include/exclude them).

### Excluding = Removing (Participants Are Gone)

When you **exclude** someone, they are **removed from your dataset entirely**.

**Example:**
```
Participant 99 did not consent
→ fury removes Participant 99
→ They do NOT appear in your final dataset
→ This cannot be undone
```

**Use excluding when:** You have strict eligibility criteria (e.g., must have consented, must be 18+).

### How to Choose?

| Situation | Use Flagging or Excluding? |
|-----------|----------------------------|
| Participant did not consent | **Exclude** (ineligible) |
| Participant is under 18 | **Exclude** (ineligible) |
| Participant failed attention check | **Flag** (might want to analyze with/without them) |
| Participant answered very quickly | **Flag** (might be valid data) |
| Participant left some questions blank | **Flag** (missing data is common) |

**Golden rule:** When in doubt, use **flagging** (you can always exclude flagged people later).

---

## 🔑 Concept 2: Pilot vs. Main Study

If you collected data in two phases (a small pilot first, then your main study), you need to tell `fury` which participants are from which phase.

**Why does this matter?**

- Reviewers want to know if you ran a pilot
- You might want to analyze pilot and main data separately
- Some journals require you to report pilot results separately

**What happens if you DON'T declare a pilot?**

- `fury` assumes ALL your data is from the main study
- `fury` will create a file that says "Pilot partition present: No"
- You can still use `fury`—this is just documenting what you did

---

## 🔑 Concept 3: Analysis-Eligible Dataset (Not "Final" Dataset)

`fury` creates what we call an **"analysis-eligible dataset per declared rules"**.

**What does that mean in plain English?**

- Your dataset has had the rules YOU declared applied to it
- People you said to exclude are excluded
- People you said to flag are flagged
- But YOU still need to make more decisions (outliers, missing data, etc.)

**Why don't we call it the "final dataset"?**

Because it's not final! You still need to:

- Decide what to do with flagged participants
- Handle missing data
- Check for outliers
- Maybe exclude more people based on your analysis

`fury` prepares your data. **You** finish it.

---

# Installing fury

## Step 1: Install Required Packages

Open R or RStudio and run these commands:

```r
# Install remotes (helps you install from GitHub)
install.packages("remotes")

# Install fury from GitHub
remotes::install_github("phdemotions/fury")

# Install haven (needed to read SPSS files)
install.packages("haven")
```

**What is GitHub?** A website where programmers share code. The `fury` package lives there.

**What is SPSS?** A statistics program. Qualtrics can export data in SPSS format, which preserves your question text and answer labels.

## Step 2: Load fury

Every time you start a new R session, you need to load `fury`:

```{r setup}
library(fury)
```

---

# Tutorial 1: Import One SPSS File (No Exclusions)

**Scenario:** You just want to import your Qualtrics data and see what variables you have.

## Step 1: Get Your SPSS File

1. Log into Qualtrics
2. Go to your survey
3. Click "Data & Analysis"
4. Click "Export & Import" → "Export Data"
5. Choose **"SPSS"** format
6. Download the file (it will be named something like `MySurvey.sav`)
7. Remember where you saved it!

## Step 2: Create a Spec File

A **"spec"** is a file that tells `fury` what to do. Think of it like a recipe.

**Option A: Using R Code (Beginner-Friendly)**

```{r create-spec-simple, eval=FALSE}
library(fury)

# Tell fury where your SPSS file is
# REPLACE THIS with your actual file path!
my_data_file <- "C:/Users/YourName/Downloads/MySurvey.sav"

# Create the spec (recipe)
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss")

# Save the spec to a file
fury_to_yaml(spec, "my_screening_rules.yaml")
```

**What just happened?**

- `fury_spec()` starts creating your recipe
- `fury_source()` tells fury where your data file is
- `fury_to_yaml()` saves your recipe to a text file

**What is YAML?** A simple text file format. It's human-readable, so you can open `my_screening_rules.yaml` in Notepad and see exactly what it says.

**Option B: Write YAML Directly (For Text File Lovers)**

Create a file called `my_screening_rules.yaml` and type this:

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"
```

**Tip:** Use forward slashes (`/`) in file paths, not backslashes (`\`).

## Step 3: Run fury

```{r run-fury-simple, eval=FALSE}
library(fury)

# Run fury on your spec file
result <- fury_run("my_screening_rules.yaml", out_dir = "my_audit")
```

**What just happened?**

- `fury` read your SPSS file
- `fury` created several output files in a folder called `my_audit`
- `result` is an object that contains information about what `fury` did

## Step 4: Look at Your Output Files

Open the folder `my_audit/audit/` on your computer. You'll see these files:

### File 1: `raw_codebook.csv`

**What it is:** A list of all the variables (questions) in your dataset.

**Why it matters:** This helps you understand what data you have.

**Open it in Excel and look for these columns:**

- `variable`: The variable name (e.g., `Q1`, `Q2`, `age`)
- `label`: The question text (e.g., "How satisfied are you?")
- `value_labels`: The answer options (e.g., "1 = Very Dissatisfied, 2 = Dissatisfied...")

**Example:**
```{r show-codebook, eval=FALSE}
# Read the codebook in R
codebook <- read.csv("my_audit/audit/raw_codebook.csv")

# Look at first few rows
head(codebook)
```

### File 2: `source_manifest.csv`

**What it is:** A record of which files `fury` imported.

**Why it matters:** If you import multiple files (pilot + main), this tracks them all.

### File 3: `session_info.txt`

**What it is:** A record of which version of R and which packages you used.

**Why it matters:** Reviewers can reproduce your exact setup.

---

# Tutorial 2: Exclude People Who Didn't Consent

**Scenario:** You want to remove anyone who clicked "I do not consent" from your dataset.

## Step 1: Find Your Consent Variable

Open your `raw_codebook.csv` and look for the consent question. Let's say it's called `consent` and the values are:

- `1 = I consent`
- `0 = I do not consent`

## Step 2: Add an Exclusion Rule

**Option A: Using R Code**

```{r exclude-consent, eval=FALSE}
library(fury)

my_data_file <- "C:/Users/YourName/Downloads/MySurvey.sav"

# Create spec with exclusion rule
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_exclude_missing("consent")  # Remove anyone with missing consent

# Save the spec
fury_to_yaml(spec, "my_screening_rules.yaml")

# Run fury
result <- fury_run("my_screening_rules.yaml", out_dir = "my_audit")
```

**What `fury_exclude_missing()` does:**

- Looks at the `consent` variable
- If someone is missing a value (they didn't answer), they are **excluded**
- Excluded people do NOT appear in your final dataset

**Option B: Write YAML Directly**

Edit your `my_screening_rules.yaml` file:

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"

  screening:
    eligibility:
      required_nonmissing: ["consent"]
      action: "exclude"
```

**What this says:** "The variable `consent` must NOT be missing. If it's missing, exclude that person."

## Step 3: Check the Results

After running `fury`, open these files:

### `screening_summary.csv`

This shows you how many people were excluded:

```
N enrolled: 100
N excluded (missing consent): 5
N analysis-eligible: 95
```

### `consort_flow.csv`

This is formatted for a **CONSORT flow diagram** (the flowchart journals often require showing how many participants you excluded and why).

**Example:**
```
stage               | n   | description
--------------------|-----|-------------------
enrolled            | 100 | Total enrolled
after_eligibility   | 95  | After excluding missing consent
analysis_eligible   | 95  | Analysis-eligible per declared rules
```

---

# Tutorial 3: Flag (But Don't Remove) Failed Attention Checks

**Scenario:** You had an attention check question where the correct answer was "3". You want to **mark** people who got it wrong, but NOT remove them yet (you'll decide later).

## Step 1: Find Your Attention Check Variable

Let's say your attention check question is called `attn_check_1` and the correct answer is `3`.

## Step 2: Add a Flag Rule

**Option A: Using R Code**

```{r flag-attention, eval=FALSE}
library(fury)

my_data_file <- "C:/Users/YourName/Downloads/MySurvey.sav"

spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_exclude_missing("consent") %>%  # Still exclude non-consenters
  fury_flag_attention(
    var = "attn_check_1",
    pass_values = 3,  # Correct answer is 3
    description = "Instructions said select 3"
  )

fury_to_yaml(spec, "my_screening_rules.yaml")

result <- fury_run("my_screening_rules.yaml", out_dir = "my_audit")
```

**What `fury_flag_attention()` does:**

- Looks at `attn_check_1`
- If the value is NOT 3, that person is **flagged**
- Flagged people STAY in your dataset (they're just marked)

**Option B: Write YAML Directly**

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"

  screening:
    eligibility:
      required_nonmissing: ["consent"]
      action: "exclude"

    quality_flags:
      attention_checks:
        - var: "attn_check_1"
          pass_values: [3]
          description: "Instructions said select 3"
          action: "flag"
```

**Notice:** `action: "flag"` (not `"exclude"`).

## Step 3: Check for Warnings

After running `fury`, open `warnings.csv`. You might see:

```
message
--------------------------------------------
7 participants flagged for attention check failures but remain in analysis pool. Review before proceeding.
```

**What this means:** 7 people failed the attention check, but they're STILL in your data. You need to decide what to do with them.

**Your options:**

1. **Keep them** (if you think the attention check wasn't important)
2. **Exclude them** (change `action: "flag"` to `action: "exclude"` and re-run)
3. **Run your analysis twice** (once with them, once without) to see if it changes your results

---

# Tutorial 4: Separate Pilot Data from Main Study

**Scenario:** You ran a pilot study in January (Jan 1-15) and then your main study in February (Feb 1-28). You want to keep them separate.

## Step 1: Find Your Date Variable

Your Qualtrics data probably has a variable called `StartDate` or `RecordedDate` that shows when each person took the survey.

## Step 2: Declare Your Pilot Partition

**Option A: Using R Code**

```{r partition-pilot, eval=FALSE}
library(fury)

my_data_file <- "C:/Users/YourName/Downloads/MySurvey.sav"

spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_partition_pilot(
    date_var = "StartDate",
    start = "2024-01-01",
    end = "2024-01-15"
  ) %>%
  fury_exclude_missing("consent")

fury_to_yaml(spec, "my_screening_rules.yaml")

result <- fury_run("my_screening_rules.yaml", out_dir = "my_audit")
```

**What `fury_partition_pilot()` does:**

- Looks at the `StartDate` variable
- Anyone who responded between Jan 1-15 is labeled as "pilot"
- Anyone who responded after Jan 15 is labeled as "main"
- **Nobody is removed**—they're just labeled

**Option B: Write YAML Directly**

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"

  screening:
    partitioning:
      pilot:
        by: "date_range"
        date_var: "StartDate"
        start: "2024-01-01"
        end: "2024-01-15"
```

## Step 3: Check the Decision Registry

Open `decision_registry.csv`. You should see:

```
decision_type           | value
------------------------|------
Pilot partition present | Yes
```

**Why this matters:** This file documents what you declared vs. what you didn't declare. Reviewers can see exactly what decisions you made.

---

# What Files Should You Share with Reviewers?

When you submit your manuscript, include these files as **supplementary materials**:

1. ✅ **`my_screening_rules.yaml`** — Shows exactly what rules you applied
2. ✅ **`decision_registry.csv`** — Shows what you declared (pilot? exclusions? flags?)
3. ✅ **`screening_summary.csv`** — Shows how many people were excluded and why
4. ✅ **`consort_flow.csv`** — Data for your participant flow diagram
5. ✅ **`source_manifest.csv`** — Shows which data files you imported
6. ✅ **`session_info.txt`** — Shows which version of R and packages you used

**How to write about this in your Methods section:**

> "Data were screened using the fury R package (Gonzales, 2026). Participants were excluded if they did not provide consent (n = 5). Participants who failed the attention check were flagged but retained for sensitivity analysis (n = 7). All screening rules were preregistered and are available in the supplementary materials (my_screening_rules.yaml). A complete audit trail, including a participant flow diagram (CONSORT format), is also available in the supplementary materials."

---

# Common Questions

## Q: Can fury compute Cronbach's alpha or check reliability?

**A: No.** `fury` only imports data and applies screening rules. You need to use other packages (like `psych`) for reliability analysis.

## Q: Can fury reverse-code my items?

**A: No.** `fury` does NOT transform your data. You'll do that in your analysis script.

## Q: Can fury run my t-test or ANOVA?

**A: No.** `fury` prepares your data. You run the statistical tests separately.

## Q: What if I realize I made a mistake in my screening rules?

**A: That's okay!** Just edit your YAML file and re-run `fury`. You'll get a new set of output files.

## Q: What's the difference between flagging and excluding again?

**A: See "Key Concepts" above.** This is THE most important thing to understand.

---

# Troubleshooting

## Problem: "Error: haven package not found"

**Solution:** Install haven:

```r
install.packages("haven")
```

## Problem: "Error: file not found"

**Solution:** Check your file path. Common mistakes:

- Using backslashes (`\`) instead of forward slashes (`/`)
- Forgetting to include the full path (e.g., just `MySurvey.sav` instead of `C:/Users/YourName/Downloads/MySurvey.sav`)
- Misspelling the file name

**Tip:** In R, you can use `file.choose()` to open a dialog box and select your file:

```r
my_file <- file.choose()
print(my_file)  # This shows you the correct path
```

## Problem: "I don't understand the output files"

**Solution:** Start with these two files (in this order):

1. **`warnings.csv`** — Tells you if something looks wrong
2. **`screening_summary.csv`** — Quick overview of what happened

If `warnings.csv` is empty, you're probably fine!

---

# Next Steps

Now that you understand `fury`, here's what to do next:

1. ✅ **Import your data** using Tutorial 1
2. ✅ **Apply screening rules** using Tutorials 2-4
3. ✅ **Check the output files** to make sure everything looks right
4. ✅ **Compute your scale scores** (in a separate R script, NOT in fury)
5. ✅ **Run your statistical analyses** (using other R packages or SPSS)
6. ✅ **Create APA tables and figures** (using other R packages)

**Remember:** `fury` is just the **first step** (data preparation). You still need to do your analysis!

---

# Summary: The Big Picture

```
┌─────────────────────────────────────────────────────────────┐
│ Step 1: Collect Data in Qualtrics                           │
│         ↓                                                    │
│ Step 2: Export as SPSS (.sav file)                          │
│         ↓                                                    │
│ Step 3: Use fury to import and screen data                  │
│         • Apply exclusion rules (eligibility)                │
│         • Apply flag rules (quality checks)                  │
│         • Separate pilot from main study                     │
│         ↓                                                    │
│ Step 4: Review fury's output files                          │
│         • warnings.csv (any problems?)                       │
│         • screening_summary.csv (how many excluded?)         │
│         • consort_flow.csv (for Methods section)             │
│         ↓                                                    │
│ Step 5: Make final decisions about flagged participants     │
│         ↓                                                    │
│ Step 6: Compute scale scores (separate R script)            │
│         ↓                                                    │
│ Step 7: Run your statistical analyses                       │
│         ↓                                                    │
│ Step 8: Create APA tables and figures                       │
│         ↓                                                    │
│ Step 9: Write your manuscript                               │
│         ↓                                                    │
│ Step 10: Share fury output files as supplementary materials │
└─────────────────────────────────────────────────────────────┘
```

**fury handles Steps 3-4. You handle the rest!**

---

# Important Reminders

## ⚠️ fury does NOT:

- Clean your data
- Compute scores
- Check reliability
- Run statistical tests
- Create a "final" dataset

## ✅ fury DOES:

- Import your SPSS file
- Apply the screening rules YOU declare
- Document what you did
- Create files for your Methods section
- Preserve an audit trail for reviewers

## 🎯 The Goal:

**Create an "analysis-eligible dataset per declared rules"** — meaning your data is ready for YOU to analyze (after you make final decisions about flagged participants, missing data, outliers, etc.).

---

# Getting Help

- 📖 **Read the main README:** `README.md` in the fury package
- 📋 **Check the examples:** `inst/examples/` folder
- 🐛 **Report bugs:** https://github.com/phdemotions/fury/issues
- 📧 **Contact the maintainer:** Josh Gonzales (see package documentation)

---

**Good luck with your research!** 🎉
