---
title: "Tutorial 1: Import SPSS Data (.sav)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 1: Import SPSS Data (.sav)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# What You'll Learn

In this tutorial, you'll learn how to:

- ✅ Import an SPSS `.sav` file from Qualtrics
- ✅ Generate a raw codebook showing your variables
- ✅ Understand what fury does (and doesn't) do with your data

**Time required:** 10 minutes

**Prerequisites:** You exported your Qualtrics survey as SPSS format

---

# Step 1: Export Your Data from Qualtrics

If you haven't already:

1. Log into Qualtrics
2. Go to your survey → "Data & Analysis"
3. Click "Export & Import" → "Export Data"
4. Choose **"SPSS"** format (not CSV!)
5. Download the file (it will be named something like `MySurvey.sav`)

**Why SPSS format?**

SPSS files preserve:
- Question text (variable labels)
- Answer option text (value labels)
- Original variable names

This makes your codebook much more readable!

---

# Step 2: Install fury (One-Time Setup)

If you haven't installed fury yet:

```{r install}
# Install remotes package (helps install from GitHub)
install.packages("remotes")

# Install fury from GitHub
remotes::install_github("phdemotions/fury")

# Install haven (required to read SPSS files)
install.packages("haven")
```

---

# Step 3: Load fury

Every time you start a new R session:

```{r setup}
library(fury)
```

**Check that haven is available:**

```{r check-haven}
if (!requireNamespace("haven", quietly = TRUE)) {
  stop("Please install haven: install.packages('haven')")
}
```

---

# Step 4: Create Your Spec File

A "spec" is a recipe that tells fury what to do. Here are two ways to create one:

## Option A: Using R Code (Recommended for Beginners)

```{r create-spec-r}
# REPLACE THIS with your actual file path!
my_data_file <- "C:/Users/YourName/Downloads/MySurvey.sav"

# Create the spec
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss")

# Save to a YAML file
fury_to_yaml(spec, "my_screening_spec.yaml")
```

**What just happened?**

- `fury_spec()` → Start building your recipe
- `fury_source()` → Tell fury where your data file is
- `fury_to_yaml()` → Save the recipe to a text file

**Tip:** Don't know your file path? Use this:

```{r find-file, eval=FALSE}
my_data_file <- file.choose()  # Opens a file picker dialog
print(my_data_file)  # Shows the full path
```

## Option B: Write YAML Directly (For Text File Lovers)

Create a file called `my_screening_spec.yaml` with:

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"
```

**Tip:** Use forward slashes (`/`) in file paths, not backslashes (`\`)!

---

# Step 5: Run fury

```{r run-fury}
# Run fury on your spec file
result <- fury_run("my_screening_spec.yaml", out_dir = "my_audit")
```

**What just happened?**

fury:
1. Read your SPSS file
2. Created several output files in the `my_audit/audit/` folder
3. Returned a `result` object with information about what it did

**How long does this take?**

- Small dataset (< 500 responses): ~5 seconds
- Medium dataset (500-5,000 responses): ~10-30 seconds
- Large dataset (> 5,000 responses): ~1 minute

---

# Step 6: Look at Your Output Files

Open the folder `my_audit/audit/` on your computer. You should see these files:

## File 1: `raw_codebook.csv` ⭐ (Look at this first!)

**What it is:** A list of all variables (questions) in your dataset.

**Open it in Excel** and you'll see columns like:

- `variable` — Variable name (e.g., `Q1`, `Q2`, `satisfaction`)
- `label` — Question text (e.g., "How satisfied are you with the product?")
- `value_labels` — Answer options (e.g., "1 = Very Dissatisfied, 2 = Dissatisfied...")

**View it in R:**

```{r view-codebook}
codebook <- read.csv("my_audit/audit/raw_codebook.csv")
head(codebook)
```

## File 2: `source_manifest.csv`

**What it is:** A record of which files fury imported and when.

**Example:**
```
source_id | file              | format | timestamp
----------|-------------------|--------|-------------------
1         | MySurvey.sav      | spss   | 2024-01-19 10:30:45
```

**Why it matters:** If you import multiple files later (pilot + main), this tracks them all.

## File 3: `session_info.txt`

**What it is:** A record of your R version and installed packages.

**Why it matters:** Reviewers can reproduce your exact setup.

---

# Understanding the Raw Codebook

Let's say your codebook looks like this:

```
variable    | label                                  | value_labels
------------|----------------------------------------|---------------------------
consent     | Do you consent to participate?         | 0=No, 1=Yes
age_group   | What is your age group?                | 1=18-24, 2=25-34, 3=35-44...
Q1          | How satisfied are you with our product?| 1=Very Dissatisfied, 2=...
Q2_1        | The product is high quality            | 1=Strongly Disagree, 2=...
```

**This tells you:**

- Variable names you'll use in your analysis (`consent`, `age_group`, `Q1`)
- What each question asked (the `label` column)
- What each number means (the `value_labels` column)

**This is NOT:**

- ❌ Computed scale scores (fury doesn't compute scores)
- ❌ Reliability estimates (fury doesn't calculate Cronbach's alpha)
- ❌ Factor loadings (fury doesn't run factor analysis)

**It's just a map of what's in your raw data.**

---

# Important: What fury Does NOT Do

```{r scope}
fury_scope()
```

fury is **only** for data preparation. It does NOT:

- ❌ Run statistical tests (t-tests, ANOVA, regression)
- ❌ Compute scale scores or reverse-code items
- ❌ Check reliability (Cronbach's alpha, factor analysis)
- ❌ Create APA-formatted tables or figures
- ❌ Clean or transform your data (no recoding, no imputation)

**fury shows you what variables you have. That's it (for now).**

In the next tutorials, you'll learn how to:
- Exclude participants who didn't consent
- Flag failed attention checks
- Separate pilot from main study data

---

# Troubleshooting

## Problem: "Error: haven package not found"

**Solution:**
```{r fix-haven, eval=FALSE}
install.packages("haven")
```

## Problem: "Error: file not found"

**Common mistakes:**

- ❌ Using backslashes: `C:\Users\...` (Windows default)
- ✅ Use forward slashes: `C:/Users/...`

**Or use `file.choose()` to let R find the path:**

```{r fix-path, eval=FALSE}
my_data_file <- file.choose()
```

## Problem: "I don't see my question text in the codebook"

**Possible causes:**

1. You exported as CSV instead of SPSS → Question text is lost
2. Your Qualtrics survey doesn't have question text set → Use variable names instead

**Solution:** Re-export as SPSS format from Qualtrics.

---

# Next Steps

Now that you've imported your data and seen the codebook, you're ready to:

📖 **[Tutorial 2: Track Pilot vs Main Study](tutorial-multi-source.html)** — Separate your pilot data

📖 **[Tutorial 3: Apply Screening Rules](tutorial-screening.html)** — Exclude non-consenters and flag attention checks

📖 **[Complete Beginner's Guide](novice-walkthrough.html)** — Full walkthrough with all concepts explained

---

# Summary

✅ **What you did:**

1. Exported Qualtrics data as SPSS
2. Created a spec file (recipe)
3. Ran `fury_run()` to import data
4. Got a raw codebook showing your variables

✅ **What you got:**

- `raw_codebook.csv` — Map of your variables
- `source_manifest.csv` — Record of what you imported
- `session_info.txt` — Reproducibility information

✅ **What fury did NOT do:**

- Did NOT compute scores
- Did NOT check reliability
- Did NOT run statistical tests
- Did NOT exclude or transform any data

**Your data is unchanged. fury just documented what's in it.**

---

# Example: Using the Included Test File

fury includes a minimal test SPSS file you can practice with:

```{r example-test-file}
library(fury)

# Use the included test file
test_file <- system.file("extdata", "test_minimal.sav", package = "fury")

# Create spec
spec <- fury_spec() %>%
  fury_source(test_file, format = "spss")

# Save and run
fury_to_yaml(spec, "test_spec.yaml")
result <- fury_run("test_spec.yaml", out_dir = tempdir())

# View codebook
codebook <- read.csv(file.path(result$artifacts$audit_dir, "raw_codebook.csv"))
print(codebook)
```

**Output:**
```
variable          label                                    value_labels
participant_id    Participant ID
condition         Experimental condition assignment        1=Control, 2=Treatment
satisfaction      How satisfied are you with the product?  1=Extremely dissatisfied, ...
```

---

**Ready for the next tutorial?** → [Tutorial 2: Track Pilot vs Main Study](tutorial-multi-source.html)
