---
title: "Tutorial 2: Track Pilot vs Main Study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 2: Track Pilot vs Main Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# What You'll Learn

In this tutorial, you'll learn how to:

- ✅ Separate pilot data from main study data
- ✅ Track provenance (what data came from where)
- ✅ Document your decisions for peer reviewers

**Time required:** 15 minutes

**Prerequisites:** You completed [Tutorial 1: Import SPSS Data](tutorial-sav-ingestion.html)

---

# Why This Matters

**Scenario:** You ran a pilot study in January to test your survey, then collected your main study data in February.

**Problem:** Both datasets are in the same Qualtrics file (or separate files you want to combine).

**Solution:** fury lets you **partition** (label) your data so pilot and main study are kept separate.

**Why peer reviewers care:**

- They want to know if you ran a pilot
- They want to see pilot results reported separately
- They want to verify you didn't change your measures between pilot and main

---

# Key Concept: Partitioning vs. Excluding

⚠️ **This is important!**

## Partitioning = Labeling (Participants Stay In)

When you **partition** data into "pilot" and "main":

- Pilot participants get a label: `partition = "pilot"`
- Main study participants get a label: `partition = "main"`
- **Nobody is removed from your dataset**

**Think of it like:** Putting sticky notes on your data that say "this is pilot" vs. "this is main"

## Excluding = Removing (Participants Are Gone)

This is different from **excluding** (which we'll cover in Tutorial 3):

- Excluded participants are **removed** from your dataset
- They don't appear in your analysis at all

**Partition first, exclude later** (if you need to exclude pilots from analysis).

---

# Step 1: Find Your Date Variable

Your Qualtrics data probably has one of these variables:

- `StartDate` — When the participant started the survey
- `RecordedDate` — When the participant finished the survey
- `EndDate` — Same as RecordedDate

**To find it:** Look at your `raw_codebook.csv` from Tutorial 1.

**Example:**
```
variable      | label
--------------|------------------------
StartDate     | Start Date
RecordedDate  | End Date
```

---

# Step 2: Decide Your Pilot Date Range

**Example scenario:**

- Pilot: January 1-15, 2024
- Main study: February 1-28, 2024

**What if I have IDs instead of dates?**

You can also partition by participant ID list (see "Advanced Options" below).

---

# Step 3: Add Pilot Partition to Your Spec

## Option A: Using R Code (Recommended)

```{r add-pilot-r}
library(fury)

# Your data file path
my_data_file <- "C:/Users/YourName/Downloads/MySurvey.sav"

# Create spec with pilot partition
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_partition_pilot(
    date_var = "StartDate",
    start = "2024-01-01",
    end = "2024-01-15"
  )

# Save to YAML
fury_to_yaml(spec, "my_screening_spec.yaml")
```

**What `fury_partition_pilot()` does:**

- Looks at the `StartDate` variable
- Anyone who responded between Jan 1-15 → labeled as "pilot"
- Everyone else → labeled as "main"

## Option B: Write YAML Directly

Edit your `my_screening_spec.yaml`:

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"

  screening:
    partitioning:
      pilot:
        by: "date_range"
        date_var: "StartDate"
        start: "2024-01-01"
        end: "2024-01-15"
```

---

# Step 4: Run fury

```{r run-fury-pilot}
result <- fury_run("my_screening_spec.yaml", out_dir = "my_audit")
```

---

# Step 5: Check the Provenance Artifacts

fury creates several files to document your partitioning decision.

## File 1: `decision_registry.csv` ⭐ (Check this first!)

**What it is:** A record of what you declared vs. didn't declare.

**Open it and look for:**

```
decision_type           | value
------------------------|------
Pilot partition present | Yes
```

**Why it matters:** This proves to reviewers that you explicitly declared a pilot partition.

**If you DON'T see "Yes":**

- You forgot to add `fury_partition_pilot()`
- Check your spec file and re-run

## File 2: `screening_summary.csv`

**What it is:** Summary statistics about your partitions.

**Example:**
```
partition | n   | description
----------|-----|-------------------
pilot     | 25  | Pilot study (Jan 1-15)
main      | 175 | Main study (after pilot)
```

## File 3: `source_manifest.csv`

**What it is:** Record of which files you imported.

**Example:**
```
source_id | file         | format | n_rows | timestamp
----------|--------------|--------|--------|-------------------
1         | MySurvey.sav | spss   | 200    | 2024-01-19 10:30:45
```

**Why it matters:** If you later combine multiple files, this tracks them all.

---

# Understanding the Output

## What Happened to Your Data?

After partitioning:

✅ **All participants are still in your dataset**

✅ **A new column was added:** `partition` (or similar)

✅ **Values in that column:**
- `"pilot"` for pilot participants
- `"main"` for main study participants

**Your data is NOT filtered or removed.**

## What If I Want to Exclude Pilots from Analysis?

**fury does NOT do this automatically.** You have two options:

### Option 1: Exclude pilots in your analysis script

```r
# Later, in your analysis script
library(dplyr)

# Filter to main study only
main_data <- my_data %>%
  filter(partition == "main")

# Run your analysis on main_data
```

### Option 2: Analyze pilot and main separately

```r
# Pilot analysis
pilot_results <- my_data %>%
  filter(partition == "pilot") %>%
  summarize(...)

# Main analysis
main_results <- my_data %>%
  filter(partition == "main") %>%
  summarize(...)
```

---

# Advanced: Multiple Partitions

You can have BOTH pilot and pretest:

```{r multiple-partitions}
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_partition_pilot(
    date_var = "StartDate",
    start = "2024-01-01",
    end = "2024-01-15"
  ) %>%
  fury_partition_pretest(
    date_var = "StartDate",
    start = "2024-01-20",
    end = "2024-01-25"
  )
```

**Result:**

- Jan 1-15 → `partition = "pilot"`
- Jan 20-25 → `partition = "pretest"`
- After Jan 25 → `partition = "main"`

---

# Advanced: Partition by ID List (Instead of Dates)

If you have a list of pilot participant IDs:

**YAML:**
```yaml
screening:
  partitioning:
    pilot:
      by: "id_list"
      id_var: "participant_id"
      ids: [1, 2, 3, 4, 5, 10, 11, 12]
```

**R Code:**
```{r partition-by-id}
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_partition_pilot(
    id_var = "participant_id",
    ids = c(1, 2, 3, 4, 5, 10, 11, 12)
  )
```

---

# Writing About This in Your Methods Section

Here's a template you can adapt:

> "Data were collected in two phases. A pilot study (N = 25) was conducted between January 1-15, 2024 to test the survey instrument. Following minor revisions to wording, the main study (N = 175) was conducted between February 1-28, 2024. Data were ingested and partitioned using the fury R package (Gonzales, 2026), which generated provenance artifacts documenting the partitioning decision (see decision_registry.csv in supplementary materials). Pilot and main study data were analyzed separately."

**Which files to share as supplementary materials:**

1. ✅ `my_screening_spec.yaml` — Shows your partitioning rules
2. ✅ `decision_registry.csv` — Proves you declared a pilot
3. ✅ `screening_summary.csv` — Shows pilot vs main counts
4. ✅ `source_manifest.csv` — Documents data provenance

---

# What If I Didn't Run a Pilot?

**That's totally fine!** Just don't call `fury_partition_pilot()`.

fury will document this:

**In `decision_registry.csv`:**
```
decision_type           | value
------------------------|------
Pilot partition present | No
```

**This is NOT an error.** It's just documenting what you did (or didn't) do.

---

# Troubleshooting

## Problem: "All my data is labeled 'main' even though I ran a pilot"

**Possible causes:**

1. Date range doesn't match your data
2. Date variable name is wrong (`RecordedDate` vs `StartDate`)
3. Date format doesn't match

**Solution:** Check your raw data to see actual dates:

```{r check-dates}
# Read your raw data
library(haven)
raw_data <- read_sav("C:/Users/YourName/Downloads/MySurvey.sav")

# Look at date range
summary(raw_data$StartDate)
```

## Problem: "I have two separate SPSS files (pilot.sav and main.sav)"

**Solution:** Import both files separately, then combine them:

```yaml
data:
  sources:
    - file: "pilot.sav"
      format: "spss"
    - file: "main.sav"
      format: "spss"
```

fury will track both in `source_manifest.csv`.

---

# Next Steps

Now that you've partitioned your data, you're ready to:

📖 **[Tutorial 3: Apply Screening Rules](tutorial-screening.html)** — Exclude non-consenters and flag attention checks

📖 **[Output Files Guide](output-files-guide.html)** — Deep dive into all fury output files

📖 **[Methods Section Guide](methods-section-guide.html)** — More templates for writing your paper

---

# Summary

✅ **What you did:**

1. Declared pilot partition by date range
2. Ran `fury_run()` to apply partitioning
3. Checked `decision_registry.csv` to verify declaration

✅ **What you got:**

- Pilot participants labeled as `partition = "pilot"`
- Main participants labeled as `partition = "main"`
- Provenance artifacts for peer review

✅ **What fury did NOT do:**

- Did NOT exclude pilot participants (they're still in your data)
- Did NOT run separate analyses (you do that later)
- Did NOT change any of your data values

**Partitioning = labeling, not removing.**

---

**Ready for the next tutorial?** → [Tutorial 3: Apply Screening Rules](tutorial-screening.html)
