---
title: "Tutorial 3: Apply Screening Rules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial 3: Apply Screening Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# What You'll Learn

In this tutorial, you'll learn how to:

- ✅ Exclude participants who didn't consent (eligibility)
- ✅ Flag (but not remove) failed attention checks (quality)
- ✅ Generate CONSORT flow diagrams for your Methods section
- ✅ Understand the critical difference between flagging and excluding

**Time required:** 20 minutes

**Prerequisites:** You completed [Tutorial 1: Import SPSS Data](tutorial-sav-ingestion.html)

---

# The #1 Mistake: Flagging vs. Excluding

⚠️ **Read this section carefully!** This is the most common source of confusion.

## Two Different Actions

fury has TWO ways to handle "bad" data:

### 1. Excluding = Removing Participants

```
action: "exclude"
→ Participants are REMOVED from your dataset
→ They do NOT appear in your final data
→ This is permanent (within fury's output)
```

**Use for:** Eligibility criteria (must consent, must be 18+, etc.)

### 2. Flagging = Marking Participants

```
action: "flag"
→ Participants are MARKED with a flag
→ They STAY in your dataset
→ You decide later whether to exclude them
```

**Use for:** Quality checks (attention checks, speeders, straight-liners)

## Decision Guide

| Situation | Action | Why |
|-----------|--------|-----|
| Did not consent | **Exclude** | Ineligible to participate |
| Under 18 years old | **Exclude** | Ineligible by IRB rules |
| Failed attention check | **Flag** | Might want to analyze with/without them |
| Completed survey very quickly | **Flag** | Might be valid data, might not |
| Has missing data on key variables | **Flag** | Missing data is common, might impute |
| Outlier on response times | **Flag** | Outliers might be real |

**Golden rule:** When in doubt, **flag** (you can always exclude later in your analysis script).

---

# Scenario 1: Exclude Non-Consenters (Eligibility)

**Goal:** Remove anyone who didn't provide consent.

## Step 1: Find Your Consent Variable

Look at your `raw_codebook.csv`. You might see:

```
variable | label                          | value_labels
---------|--------------------------------|------------------
consent  | Do you consent to participate? | 0=No, 1=Yes
```

## Step 2: Add Exclusion Rule

**Option A: Using R Code**

```{r exclude-consent}
library(fury)

my_data_file <- "C:/Users/YourName/Downloads/MySurvey.sav"

spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_exclude_missing("consent")  # Remove if consent is missing

fury_to_yaml(spec, "my_screening_spec.yaml")
```

**What `fury_exclude_missing()` does:**

- Looks at the `consent` variable
- If someone has a missing value (NA, blank, or didn't answer) → **excluded**
- Excluded people do NOT appear in your output data

**Option B: Write YAML Directly**

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"

  screening:
    eligibility:
      required_nonmissing: ["consent"]
      action: "exclude"
```

## Step 3: Run fury

```{r run-exclude}
result <- fury_run("my_screening_spec.yaml", out_dir = "my_audit")
```

## Step 4: Check the Results

### `screening_summary.csv`

```
stage                | n   | description
---------------------|-----|------------------------
enrolled             | 200 | Total enrolled
after_eligibility    | 195 | After excluding missing consent
analysis_eligible    | 195 | Analysis-eligible per declared rules
```

**Interpretation:** 5 people were excluded for missing consent.

### `consort_flow.csv`

This file is formatted for creating a **CONSORT flow diagram** (the flowchart journals require):

```
stage                | n   | exclusion_reason       | excluded_n
---------------------|-----|------------------------|------------
enrolled             | 200 | -                      | -
after_eligibility    | 195 | Missing: consent       | 5
analysis_eligible    | 195 | -                      | -
```

**Use this to create your flowchart in your paper.**

---

# Scenario 2: Flag Failed Attention Checks (Quality)

**Goal:** Mark (but don't remove) people who failed an attention check.

## Step 1: Find Your Attention Check Variable

Let's say your survey had this question:

> "To ensure you are paying attention, please select 'Somewhat agree' for this question."

In your codebook:

```
variable      | label                  | value_labels
--------------|------------------------|--------------------------------
attn_check_1  | Attention check item   | 1=Strongly disagree, 2=Disagree,
              |                        | 3=Somewhat disagree, 4=Somewhat agree,
              |                        | 5=Agree, 6=Strongly agree
```

**Correct answer:** 4 (Somewhat agree)

## Step 2: Add Flag Rule

**Option A: Using R Code**

```{r flag-attention}
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_exclude_missing("consent") %>%  # Still exclude non-consenters
  fury_flag_attention(
    var = "attn_check_1",
    pass_values = 4,  # Correct answer
    description = "Instructions said select 'Somewhat agree'"
  )

fury_to_yaml(spec, "my_screening_spec.yaml")
```

**What `fury_flag_attention()` does:**

- Looks at `attn_check_1`
- If the value is NOT 4 → person is **flagged**
- Flagged people STAY in your dataset (they're just marked)

**Option B: Write YAML Directly**

```yaml
data:
  sources:
    - file: "C:/Users/YourName/Downloads/MySurvey.sav"
      format: "spss"

  screening:
    eligibility:
      required_nonmissing: ["consent"]
      action: "exclude"

    quality_flags:
      attention_checks:
        - var: "attn_check_1"
          pass_values: [4]
          description: "Instructions said select 'Somewhat agree'"
          action: "flag"
```

## Step 3: Run fury

```{r run-flag}
result <- fury_run("my_screening_spec.yaml", out_dir = "my_audit")
```

## Step 4: Check for Warnings

### `warnings.csv` ⚠️

```
severity | message
---------|------------------------------------------------------------
warning  | 12 participants flagged for attention check failures but
         | remain in analysis pool. Review before proceeding.
```

**What this means:**

- 12 people failed the attention check
- They are STILL in your data (flagged, not excluded)
- You need to decide what to do with them

**Your options:**

### Option 1: Keep Them (No Additional Action)

If you think the attention check wasn't important, just proceed. They're in your data.

### Option 2: Exclude Them (Change action to "exclude")

Edit your spec to change `action: "flag"` to `action: "exclude"`:

```yaml
quality_flags:
  attention_checks:
    - var: "attn_check_1"
      pass_values: [4]
      description: "Instructions said select 'Somewhat agree'"
      action: "exclude"  # Changed from "flag"
```

Re-run fury.

### Option 3: Run Analysis Twice (Sensitivity Analysis)

```r
# Analysis 1: Include everyone
all_data <- read.csv("my_audit/data/analysis_eligible_data.csv")
results_all <- analyze(all_data)

# Analysis 2: Exclude flagged participants
library(dplyr)
no_flags <- all_data %>%
  filter(attn_check_1_flag == FALSE)  # Assuming fury added this flag column
results_no_flags <- analyze(no_flags)

# Compare results
compare(results_all, results_no_flags)
```

---

# Combining Multiple Rules

You can have BOTH eligibility exclusions AND quality flags:

```{r combine-rules}
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%

  # Eligibility (exclude)
  fury_exclude_missing(c("consent", "age")) %>%

  # Quality flags (don't exclude)
  fury_flag_attention(
    var = "attn_check_1",
    pass_values = 4,
    description = "Instructions said select 'Somewhat agree'"
  ) %>%
  fury_flag_attention(
    var = "attn_check_2",
    pass_values = 2,
    description = "Instructions said select 'Disagree'"
  )

fury_to_yaml(spec, "my_screening_spec.yaml")
result <- fury_run("my_screening_spec.yaml", out_dir = "my_audit")
```

**Result:**

- People missing consent or age → **excluded** (removed)
- People who failed attn_check_1 or attn_check_2 → **flagged** (marked, but kept)

---

# Understanding the CONSORT Flow

After running screening rules, `consort_flow.csv` shows the participant flow:

```
stage                | n   | exclusion_reason           | excluded_n
---------------------|-----|----------------------------|------------
enrolled             | 200 | -                          | -
after_eligibility    | 190 | Missing: consent, age      | 10
after_quality        | 190 | Flagged: attn_check_1 (12) | 0
analysis_eligible    | 190 | -                          | -
```

**Key points:**

- N enrolled: 200
- N excluded for eligibility: 10 (missing consent or age)
- N flagged for quality: 12 (failed attention check, but NOT excluded)
- N analysis-eligible: 190

**Notice:** Flagged participants are NOT excluded (excluded_n = 0 for quality stage).

---

# Writing About This in Your Methods Section

Here's a template:

> "Data were screened using the fury R package (Gonzales, 2026). Participants were excluded if they did not provide consent or were missing age information (n = 10). Participants who failed at least one attention check were flagged but retained for sensitivity analysis (n = 12). The final analysis-eligible sample per declared rules consisted of N = 190 participants. All screening rules were preregistered (see my_screening_spec.yaml in supplementary materials). A complete CONSORT flow diagram is available in the supplementary materials (consort_flow.csv)."

**Sensitivity analysis statement (if you ran analysis twice):**

> "Results were substantively identical when excluding flagged participants (see Appendix A for sensitivity analysis)."

---

# Advanced: Multiple Attention Checks

If you have multiple attention checks and want to flag people who failed ANY of them:

```{r multiple-checks}
spec <- fury_spec() %>%
  fury_source(my_data_file, format = "spss") %>%
  fury_flag_attention(var = "attn_check_1", pass_values = 4,
                      description = "Select 'Somewhat agree'") %>%
  fury_flag_attention(var = "attn_check_2", pass_values = 2,
                      description = "Select 'Disagree'") %>%
  fury_flag_attention(var = "attn_check_3", pass_values = 1,
                      description = "Select 'Strongly disagree'")
```

fury will create separate flags for each check. You can later decide:

- Exclude if they failed **ANY** check
- Exclude if they failed **ALL** checks
- Exclude if they failed **2 or more** checks

---

# Advanced: Custom Exclusion Logic

For more complex rules, use YAML directly:

```yaml
screening:
  eligibility:
    custom_rules:
      - predicate: "age >= 18"
        description: "Must be 18 or older"
        action: "exclude"
      - predicate: "consent == 1"
        description: "Must have consented"
        action: "exclude"
```

---

# Troubleshooting

## Problem: "Everyone is flagged!"

**Possible causes:**

1. Wrong pass value (you said 4, but correct answer was actually 3)
2. Wrong variable name
3. Value labels don't match actual values

**Solution:** Check your raw data:

```{r check-values}
library(haven)
raw_data <- read_sav("C:/Users/YourName/Downloads/MySurvey.sav")
table(raw_data$attn_check_1)  # See actual values
```

## Problem: "I want to exclude flagged participants, but I already ran fury"

**Solution:** Two options:

### Option 1: Re-run fury with action: "exclude"

Change your spec and re-run. This is the cleanest approach.

### Option 2: Filter in your analysis script

```r
library(dplyr)

# Assuming fury added flag columns
clean_data <- my_data %>%
  filter(!attn_check_1_flag)  # Remove flagged participants
```

---

# What Files to Share with Reviewers

When you submit your paper, include these as supplementary materials:

1. ✅ `my_screening_spec.yaml` — Your declared screening rules
2. ✅ `decision_registry.csv` — What you declared vs didn't declare
3. ✅ `screening_summary.csv` — Quick overview of exclusions/flags
4. ✅ `consort_flow.csv` — Data for your CONSORT flow diagram
5. ✅ `warnings.csv` — Any warnings fury generated
6. ✅ `source_manifest.csv` — Which files you imported

**Reviewers can:**

- See exactly what rules you applied
- Verify you followed your preregistration
- Reproduce your screening process

---

# Next Steps

Now that you've applied screening rules, you're ready to:

📖 **[Output Files Guide](output-files-guide.html)** — Detailed explanation of all fury output files

📖 **[Methods Section Guide](methods-section-guide.html)** — More templates and examples

📖 **[Complete Beginner's Guide](novice-walkthrough.html)** — Full walkthrough with all concepts

---

# Summary

✅ **What you did:**

1. Excluded participants for eligibility (consent, age)
2. Flagged participants for quality (attention checks)
3. Generated CONSORT flow artifacts

✅ **What you got:**

- `screening_summary.csv` — Overview of exclusions/flags
- `consort_flow.csv` — Data for your Methods section flowchart
- `warnings.csv` — Alerts about flagged participants
- Analysis-eligible dataset per declared rules

✅ **Critical distinction:**

- **Exclude** = Remove (for eligibility criteria)
- **Flag** = Mark (for quality checks, decide later)

⚠️ **Remember:**

fury produces an **"analysis-eligible dataset per declared rules"**

This is NOT a "final" or "cleaned" dataset. You still need to:
- Decide what to do with flagged participants
- Handle missing data
- Check for outliers
- Make other analysis decisions

**fury prepares your data. YOU finish it.**

---

**Ready to learn more?** → [Output Files Guide](output-files-guide.html)
